---
title: "plotly and Mapview"
output: 
  html_document:
    keep_md: true
self_contained: yes
---

<script src="../../../js/hideoutput.js"></script>

```{r echo = F, message = F, warning = F}
knitr::opts_chunk$set(fig.path = '../../../img/', message = F, warning = F)

library(tidyverse)
library(maps)
library(sf)
library(raster)
library(tmap)
library(micromap)
library(ggrepel)
library(plotly)
```

We've seen how to use some of the basic plotting features in R to create static maps in the last two sub-modules.  Although static maps are still the de facto format for publications, interact with the plot data can be much more informative.  R is no longer limited to static plotting since several new package have been introduced in the last few years.  Most of these packages link R code to existing plotting libraries that were developed in other languages (e.g., JavaScript).  These are not new packages in the computing world, just new to the R universe.  They are insanely easy to use if you know just a bit of R code and they interface really well with R Markdown.  

We'll introduce the following packages in this submodule:

* [plotly](https://plot.ly/r/): interactive graphs in R, not limited to maps
* [mapview](https://github.com/r-spatial/mapview): quick interactive maps to examine eand visually investigate geometries and attributes of spatial data

<!-- https://moderndata.plot.ly/visualizing-geo-spatial-data-with-sf-and-plotly/ -->

### plotly

The plotly library lets you easily create interactive plots in R.

```{r}
library(plotly)
p <- plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length)
p
```

### mapview

The mapview library lets you easily create interactive maps in R.

```{r}
library(mapview)
mapview(breweries)
```
